Exploring how changes in cause of death have narrowed the black-white life

### 2017-02-24 Manuscript/Appendix-agestd-mort-rates.Rmd

# Create unique blank strip labels for empty facets
bl = sapply(1:37, function(n) paste(rep(" ",n),collapse=""))

mortality.rates$state.reorder2 <- factor(mortality.rates$state,
                                        levels = c(bl[1:10], "Maine",
                                                   bl[11:19], "Vermont", "New Hampshire",
                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
                                                   bl[30:33], "Texas", bl[34:37], "Florida"))

mortality.rates.diff$state.reorder2 <- factor(mortality.rates.diff$state,
                                        levels = c(bl[1:10], "Maine",
                                                   bl[11:19], "Vermont", "New Hampshire",
                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
                                                   bl[30:33], "Texas", bl[34:37], "Florida"))
```

```{r plot-arrange-as-us-map, echo=F, fig.width=18.5, fig.height=14}

# sex_i <- "Male"
# cod <- "Injuries"

# for(sex_i in c("Male")) {
#   for(cod in levels(mortality.rates$COD)[1]) {
    data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
    map.plot1 <- ggplot(data.sub,
                        aes(y = rate.per.100k_mean, x = year)) +
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race), fill = "grey", col = NA) +
      geom_line(aes(lty = Race)) +
      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
      theme_classic(base_size = 20) +
      theme(axis.text.x = element_blank(),
            strip.background=element_blank(),
            axis.line=element_blank(),
            axis.ticks=element_blank()) +
      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Age-standardized mortality rate (per 100,000)") +
      xlab("Year (1969-2013)") +
      ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod))# + theme_minimal()

    print(map.plot1)

        map.plot2 <- ggplot(subset(mortality.rates.diff, sex == sex_i & COD == cod),
                        aes(y = rate.difference_mean, x = year)) +
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), fill = "grey", col = NA) +
      geom_line() +
      geom_hline(aes(yintercept = 0), col = "#636363") +
      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
      theme_classic(base_size = 20) +
      theme(axis.text.x = element_blank(),
            strip.background=element_blank(),
            axis.line=element_blank(),
            axis.ticks=element_blank()) +
      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Black-white difference (per 100,000)") +
      xlab("Year (1969-2013)") +
      ggtitle(paste0("Black-white difference in age-adjusted mortality (per 100,000) in ", 
                     sex_i, " for ", cod))

    print(map.plot2)

>>>>

# Create unique blank strip labels for empty facets
bl = sapply(1:38, function(n) paste(rep(" ",n),collapse=""))

mortality.rates$state.reorder2 <- factor(mortality.rates$state,
                                        levels = c(bl[1:10], "Maine",
                                                   bl[11:19], "Vermont", "New Hampshire",
                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
                                                   bl[30:33], "Texas", bl[34:38], "Florida"))

mortality.rates.diff$state.reorder2 <- factor(mortality.rates.diff$state,
                                        levels = c(bl[1:10], "Maine",
                                                   bl[11:19], "Vermont", "New Hampshire",
                                                   "Washington", "Idaho", "Montana", "North Dakota", "Minnesota", "Illinois", "Wisconsin", "Michigan", "New York", "Massachusetts", "Rhode Island",
                                                   "Oregon", "Nevada", "Wyoming", "South Dakota", "Iowa", "Indiana", "Ohio", "Pennsylvania", "New Jersey", "Connecticut", bl[20],
                                                   "California", "Utah", "Colorado", "Nebraska", "Missouri", "Kentucky", "West Virginia", "Virginia", "Maryland", "Washington DC", bl[21],
                                                   bl[22], "Arizona", "New Mexico", "Kansas", "Arkansas", "Tennessee", "North Carolina", "South Carolina", "Delaware", bl[23:24],
                                                   bl[25:27], "Oklahoma", "Louisiana", "Mississippi", "Alabama", "Georgia", bl[28:29],
                                                   bl[30:33], "Texas", bl[34:38], "Florida"))
```

```{r plot-arrange-as-us-map, echo=F, fig.width=19.5, fig.height=14, results="asis", message = F}
roundUp <- function(x) 10^ceiling(log10(x))

# sex_i <- "Male"
# cod <- "Cardiovascular"

for(sex_i in c("Male", "Female")) {
  for(cod in levels(mortality.rates$COD)[c(3, 2, 4, 6, 5, 1)]) {
# for(sex_i in c("Male")) {
#   for(cod in levels(mortality.rates$COD)[1]) {
    cat("  \n### Age-adjusted mortality (per 100,000) in ", sex_i, " for ", cod, "  \n")

    data.sub <- subset(mortality.rates, sex == sex_i & COD == cod)
    lim.y <- max(data.sub$rate.per.100k_mean)
    map.plot1 <- ggplot(data.sub,
                        aes(y = rate.per.100k_mean, x = year)) +
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_line(aes(lty = Race, col = Race)) +
      geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race, fill = Race), col = NA, alpha = 0.5) +
      scale_color_manual(values = c("#2166ac", "#b2182b")) +
      scale_fill_manual(values = c("#67a9cf", "#ef8a62")) +
      scale_y_continuous(limits = c(0, lim.y)) + 
      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
      theme_classic(base_size = 20) +
      theme(axis.text.x = element_blank(),
            strip.background=element_blank(),
            axis.line=element_blank(),
            axis.ticks=element_blank()) +
      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Age-standardized mortality rate (per 100,000)") +
      xlab("Year (1969-2013)")
    print(map.plot1)

    cat("  \n")

    cat("  \n### Black-white difference in age-adjusted mortality (per 100,000) in ", 
        sex_i, " for ", cod, "  \n")

    data.sub2 <- subset(mortality.rates.diff, sex == sex_i & COD == cod)
    lim.y2 <- max(data.sub2$rate.difference_mean)

    map.plot2 <- ggplot(data.sub2,
                        aes(y = rate.difference_mean, x = year)) +
      geom_vline(aes(xintercept = 1983), col = "#636363", lty = 3) +
      geom_vline(aes(xintercept = 1993), col = "#636363", lty = 3) +
      geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), fill = "grey", col = NA) +
      geom_line() +
      geom_hline(aes(yintercept = 0), col = "#636363") +
      scale_y_continuous(limits = c(0, lim.y)) + 
      facet_wrap( ~ state.reorder2, ncol = 11, drop = F, strip.position="bottom") +
      theme_classic(base_size = 20) +
      theme(axis.text.x = element_blank(),
            strip.background=element_blank(),
            axis.line=element_blank(),
            axis.ticks=element_blank()) +
      #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      ylab("Black-white difference (per 100,000)") +
      xlab("Year (1969-2013)")

    print(map.plot2)
    cat("  \n")
    cat("  \n")
  }
  cat("  \n")
}
```

### 2017-02-24 Manuscript/Manuscript_COD_decomp.Rmd

---
title: Exploring how changes in cause of death have narrowed the black-white life expectancy gap in the United States.

output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

@@

  max.value <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
                                            cod.df$era == "1969-1983" &
                                            cod.df$sex == "Male" ])

  max.cause <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-j, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "1969-1983" & 
                    cod.df$Contribution.to.change == max.value]

  min.value <- min(cod.df$Contribution.to.change[cod.df$state == diff[i, ]$state & 
                                            cod.df$era == "1969-1983" &
                                            cod.df$sex == "Male" ])

  min.cause <- cod.df$COD[cod.df$state == diff[i, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "1969-1983" & 
                    cod.df$Contribution.to.change == min.value]

@@

  min.value.DC <- min(cod.df$Contribution.to.change[cod.df$state == "Washington DC" & 
                                            cod.df$era == "1983-1993" &
                                            cod.df$sex == "Male" ])

  min.cause.DC <- cod.df$COD[cod.df$state == "Washington DC" &
                    cod.df$sex == "Male" &
                    cod.df$era == "1983-1993" & 
                    cod.df$Contribution.to.change == min.value.DC]
```

@@

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `. By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `. During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

@@

```{r figure2, echo = F, fig.height=9, fig.width=11}
# - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.

@@

```{r figure3, echo = F, fig.height=9, fig.width=11}
y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))

@@

```{r, fig.align="center", fig.height=5, fig.width=9, echo=F, eval = F}
males.era2.2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "era2"), aes(y = state.gdiff.num.1, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era2.2

@@

males.era3.2

@@

```{r, echo =F}
sum5 <- BlackWhite_results %>% filter(year == 1969 & sex == "Female") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
```

@@

```{r figure4, fig.align = "center", echo=F}

@@

```{r figure2b, fig.align = "center", echo=F}
y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))

@@

```{r fig6, fig.align = "center", echo=F}
y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))

@@

```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

@@

**Pennsylvania**
```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

pa.1993 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
pa.1983 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")

@@

**Connecticut**
```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

ct.1993 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
ct.1983 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")

@@

**Mississippi**
```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

ms.1993 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
ms.1983 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")

@@

**Florida**
```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9, eval=F}
fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs))

fl.1993 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")
fl.1983 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs",
                                  decomp.var.prop = "age_COD_cont_prop", type.of.data = "practice")

@@

```{r}
cod_change_results2 <- data.frame(cod_change_results[c(3, 40, 41)], apply(cod_change_results[4:39], 2, function(x) x*(-1)))

@@

data.table::data.table(cod_change_results2 %>% filter(Sex == "Male"))
```

@@

```{r}
data.table::data.table(cod_change_results2 %>% filter(Sex == "Female"))
```

>>>>

---
title: Exploring how changes in cause of death have narrowed the black-white life
  expectancy gap in the United States.
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

@@

  max.value <- max(cod.df$Cx_diff_mean[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
                                            cod.df$era == "era1" &
                                            cod.df$sex == "Male" ])

  max.cause <- cod.df$Cause.of.death[cod.df$state == diff[dim(diff)[1]-j, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "era1" & 
                    cod.df$Cx_diff_mean == max.value]

  min.value <- min(cod.df$Cx_diff_mean[cod.df$state == diff[i, ]$state & 
                                            cod.df$era == "era1" &
                                            cod.df$sex == "Male" ])

  min.cause <- cod.df$Cause.of.death[cod.df$state == diff[i, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "era1" & 
                    cod.df$Cx_diff_mean == min.value]

@@

  min.value.DC <- min(cod.df$Cx_diff_mean[cod.df$state == "Washington DC" & 
                                            cod.df$era == "era2" &
                                            cod.df$sex == "Male" ])

  min.cause.DC <- cod.df$Cause.of.death[cod.df$state == "Washington DC" &
                    cod.df$sex == "Male" &
                    cod.df$era == "era2" & 
                    cod.df$Cx_diff_mean == min.value.DC]
```

@@

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `. By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `. During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

@@

```{r figure2, echo = F, fig.height=9, fig.width=10}
# - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.

@@

```{r figure3, echo = F, fig.height = 9, fig.width = 11}
y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))

@@

**All the above together as one figure**

```{r, fig.align="center", fig.height=9, fig.width=15, echo=F, eval = T}
males.era2.2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "era2"), aes(y = state.gdiff.num.1, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

@@

@@

```{r, echo =F}
sum5 <- BlackWhite_results %>% filter(year == 1969 & sex == "Female") %>% 
  arrange(LE_wbgap_mean) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
```

@@

```{r figure2a, fig.align = "center", echo=F, fig.height = 9, fig.width = 7.5}

@@

```{r figure2b, fig.align = "center", echo=F, fig.height = 9, fig.width = 7.5}
y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))

@@

```{r fig6, fig.align = "center", echo=F, fig.height = 9, fig.width = 9}
y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))

@@

```{r ny_1969_1983_age_cod_decomp, echo = F, fig.width = 9}
ny.dat <- age_cod_results %>% filter(state == "New York" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))

@@

**Pennsylvania**
```{r pa_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
pa.dat <- age_cod_results %>% filter(state == "Pennsylvania" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))

pa.1993 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
pa.1983 <- make_pretty_decomp_plot(decomp.table = subset(pa.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")

@@

**Connecticut**
```{r ct_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
ct.dat <- age_cod_results %>% filter(state == "Connecticut" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))

ct.1993 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
ct.1983 <- make_pretty_decomp_plot(decomp.table = subset(ct.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")

@@

**Mississippi**
```{r ms_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
ms.dat <- age_cod_results %>% filter(state == "Mississippi" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))

ms.1993 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
ms.1983 <- make_pretty_decomp_plot(decomp.table = subset(ms.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")

@@

**Florida**
```{r fl_1993_1983_age_cod_decomp, echo = F, fig.width = 9}
fl.dat <- age_cod_results %>% filter(state == "Florida" & year %in% c(1969, 1983, 1993) & sex == "Male") %>% 
                                     mutate(sign.var = sign(age_COD_cont_yrs_mean))

fl.1993 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1993), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")
fl.1983 <- make_pretty_decomp_plot(decomp.table = subset(fl.dat, year == 1983), sign.var = "sign.var",
                                  strat.var.1 = "age_minbin", partition.bar.var = "COD", decomp.var = "age_COD_cont_yrs_mean",
                                  decomp.var.prop = "age_COD_cont_prop_mean", type.of.data = "practice")

@@

```{r, echo=F}
cod_change_results2 <- data.frame(cod_change_results[c(3, 40, 41)], apply(cod_change_results[4:39], 2, function(x) x*(-1)))

@@

DT::datatable(cod_change_results2 %>% filter(Sex == "Male"))
```

@@

```{r, echo=F}
DT::datatable(cod_change_results2 %>% filter(Sex == "Female"))
```
